---
title: "Prepare peak data"
author: "ajpelu"
date: "2023-05-31"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(janitor) 
library(lubridate)
library(ggpubr)
```


```{r}
# Read the transect data 
transectos <- read_csv(here::here("data/transectos_metadata.csv"))
```


### Get the peak day of the phenology by year and site for each species 

- JMBA give the a folder with files for each species of the curve fitted by year and site (`/raw_data/pheno_sp`)
- We prepare the data: read the data, join the transects metadata and then compute the day of year with the maximum fitted valued (peak_day) 
- There are problems with the files: `Glaucopsyche_alexis_BMSSN_pheno.xlsx` and `Libythea_celtis_BMSSN_pheno.xlsx`. We removed them into folder (`/raw_data/pheno_sp_novalid`)

- I wrote a custom function to read the data, unite year, month and day and convert it into date. Then compute the day of year (`doy`), generate a dataframe (`pheno`) and export the data as `/data/pheno.csv`

```{r}
# List the files 
pheno_files <- list.files(here::here("data/raw_data/pheno_sp/"), pattern = "*pheno.xlsx", full.names = TRUE)


prepare_pheno <- function(file) { 
  df_pheno <- readxl::read_excel(file) |>
    janitor::clean_names() |> 
    mutate(date = lubridate::make_date(year, month, day)) |>
    mutate(doy = lubridate::yday(date)) 
  
  return(df_pheno)
  }


pheno <- purrr::map_df(pheno_files, ~ prepare_pheno(.)) 

write_csv(pheno, here::here("data/pheno.csv"))
```

- The dataframe `pheno` has `r nrow(pheno)` records.  

- We extract the doy (day of the year) for each species, sites and year with the maximum peak of the fitted valued. For this purpose we write a custom function `get_peak_day`. The dataframe `pheno` contains some records with `NA` value in `site_id` and also some data with `NA` values in `fitted` variable. So we remove them using the option `na.rm` in the function `dplyr::slice_max`. The information about transect was joined. The output results is exported as `data/doy_peak_sps.csv`) 

```{r}
get_peak_day <- function(df_pheno) { 
  peak_day <- df_pheno |> 
    group_by(species, site_id, year) |>
    slice_max(fitted, na_rm = TRUE) |> 
    dplyr::select(species, site_id, year,
                month, day, week,
                fitted, date, doy) |> 
  ungroup()
  
  return(peak_day)
  }
 
doy_peak_sps <- get_peak_day(pheno) |> 
  inner_join(transectos)

write_csv(doy_peak_sps, here::here("data/doy_peak_sps.csv"))
```


